Sound Similarity, Sound Matching, and Differentiable Audio Synthesizer Control

### 2025-04-29 main.tex
Updates from Overleaf

\begin{abstract}
Manual sound design with a synthesizer is an iterative process. When an artist has a target sound in mind, they listen to the output of their chosen synthesizer, determine whether the output is closer or farther from the target, and adjust the synthesizer's parameters until satisfied. There are practically an infinite number of parameter combinations, and their effect on the output sound is hard to predict. Iterative sound-matching is the automation of this process, more formally, the automatic programming of an audio synthesizer to generate outputs similar to a target sound while guided by a loss function that quantifies sound similarity.
Past works in sound-matching often relied on a narrow range of synthesizers and overlooked the possible interactions between the loss function and synthesizer parameters.
We analyzed the interplay between four simple subtractive, additive, and FM synthesizers when guided by four different novel or established similarity functions.
For each loss function and synthesizer pair, we ran 300 experiments with randomized initial values and targets, and used the final parameter configurations and spectrogram distances as key performance evaluation metrics of the performance. A post-hoc analysis of these performance measures showed that the best‐performing loss function varies based on the method of synthesis at hand, which highlights the value of novel similarity measures and de-prioritizes the search for the elusive universally optimal sound similarity measure.
\end{abstract}

@@

\section{Introduction}
\subsection{Digital Synthesis With Sound-Matching}
Since the 1970s, artists such as Suzanne Ciani~\cite{ciani_life_in_waves} have explored the use of audio synthesizers to replicate natural sounds—such as a bottle cap popping off—in a way that preserves the essential qualities of the original while showcasing the characteristics of the synthesizer. This approach goes beyond mere replication and uses synthesizers for artistic re-interpretation of sounds. Pursuing this idea, we aim to emulate and automate this process using digital audio synthesizers.

@@

Automatic sound design has been a subject of research for decades, with reduction of tinkering time and creation of ``interesting" sounds as the main motivators \cite{krekovic2019insights,turian2020sorry,horner1993machine,salimi2020make,esling2019flow,engel2020ddsp,mitchell2007evolutionary,shier2020spiegelib}. \textbf{Sound-matching} is a common form of automatic sound design explored in literature, where the goal is to find parameters for a synthesizer that replicate \textit{all} or \textit{some} of the characteristics of a target sound as best as possible, given some kind of similarity metric~\cite{horner1993machine,mitchell2007evolutionary}. Not all sound-matching works follow the iterative process of manual sound design. \textit{Iterative} sound-matching refers to works which mimic the back and forth process of listening to a synthesizer's output and updating the parameters until the termination conditions are met.

A brief description of the iterative sound-matching workflow may be useful here: The two necessary components for sound-matching are the \textit{loss function}, which determines how close two sounds are, and the \textit{parametric synthesizer} that outputs sounds based on input parameters. In the iterative workflow, a target sound is selected, the synthesizer outputs sounds based on its parameters, and the loss function determines the proximity of the output to the target. From there, the synthesizer parameters are updated, the output and target are compared, and so on.

With the goal of implementing effective iterative sound-matching algorithms for sound designers, we identify a number of weaknesses in current relevant literature (see Section~\ref{sec:lacking}). The experiments in this work pertain to two core issues, which we named \LossSelect~and \SynthSelect. The former refers to the search for an optimal loss function, lack of novel algorithms, and the focus on outperforming state of the art (\gls{SOTA}), while the latter refers to the lack of diversity in synthesis methods, and their interaction with the loss function.

@@

where $T$ represents time in seconds, and $n$ is the frequency parameter. If $n$ is 1, then $x(n)$ would be a waveform with a frequency of 1 hertz, meaning that the value of $x(n)$ oscillates to the same point once every second. Waveform generators are called oscillators for this reason. If $n$ is in the range of 30 to 14,000 (the upper limit can be higher for some), $x(n)$ would be an audible tone~\cite{smith2007mathematics,jilek2014reference}.

@@

Differential DSP (\gls{DDSP}) is a general term for applications that combine machine learning techniques such as gradient descent~\cite{goodfellow2016deep,boyd2004convex} with DSP~\cite{engel2020ddsp}. Here, we take a differential approach to sound-matching using DDSP, i.e., the use of differentiable loss functions and gradient descent to update synthesizer parameters to reduce the difference between the output and the target sound.

Implementing complex DSP functions in a differentiable manner can be challenging, and effective differentiable audio similarity measures require careful mathematical expression of the desired attributes of sound. These issues have likely contributed to the limited exploration of this domain. Perhaps the most relevant recent work is the comparison of two differentiable loss functions with a simple FM synthesizer by Vahidi \textit{et al.}~\cite{vahidi2023mesostructures}.

@@

In sound-matching, particularly in a Deep Learning (\gls{DL}) context~\cite{goodfellow2016deep}, the similarity function can also be called a \textit{loss} function, where the emphasis is on measurement and reduction of the distance between target and output. It is important to note that there is a close relationship between the loss function $L$ and the sound representation function $\phi$. A distance measure $d$ is applied to the features extracted by $\phi$.

@@

\subsection{Automatic Sound Matching}
\label{sec:sound_matching_definition}
In past works the sound-matching problem has been tackled from a variety of perspectives as it involves a number of modular parts that need to be considered: the choice of synthesizer $g$, the target sounds of interest, the representation function $\phi$, and the heuristic for finding the optimal $\theta$. We discuss these components in the upcoming sections, and give a coherent typology of the sound-matching tasks in Section~\ref{sec:matching_types}.

>>>>

\begin{abstract}
Manual sound design with a synthesizer is an iterative process. When an artist has a target sound in mind, they listen to the output of their chosen synthesizer, determine whether the output is closer or farther from the target, and adjust the parameters until satisfied. There are practically an infinite number of parameter combinations, and their effect on the output sound is hard to predict. Iterative sound-matching is the automation of this process, more formally, the automatic programming of an audio synthesizer to generate outputs similar to a target sound while guided by a loss function that quantifies sound similarity.
Past works in sound-matching often relied on a narrow range of synthesizers and overlooked the possible interactions between the loss function and synthesizer parameters.
We analyzed the interplay between four simple subtractive, additive, and FM synthesizers when guided by four different novel or established similarity functions.
For each loss function and synthesizer pair, we ran 300 experiments with randomized initial values and targets, and used the final parameter configurations and spectrogram distances as key performance evaluation metrics of the performance. A post-hoc analysis of these performance measures showed that the best‐performing loss function varies based on the method of synthesis at hand, which highlights the value of novel similarity measures and de-prioritizes the search for the elusive universally optimal sound similarity measure.
\end{abstract}

@@

\section{Introduction}
\subsection{Digital Synthesis With Sound-Matching}
Since the 1970s, artists such as Suzanne Ciani~\cite{ciani_life_in_waves} have explored the use of audio synthesizers to replicate natural sounds—such as a bottle cap popping off—in a way that preserves the essential qualities of the original while showcasing the characteristics of the synthesizer. This approach goes beyond mere replication and uses synthesizers for artistic re-interpretation of sounds. Pursuing this idea, we aim to automate this process using digital audio synthesizers.

@@

Automatic sound design has been a subject of research for decades, with reduction of tinkering time and creation of ``interesting" sounds as the main motivators \cite{krekovic2019insights,turian2020sorry,horner1993machine,salimi2020make,esling2019flow,engel2020ddsp,mitchell2007evolutionary,shier2020spiegelib}. \textbf{Sound-matching} is a common form of automatic sound design explored in literature, where the goal is to find parameters for a synthesizer that replicate \textit{all} or \textit{some} of the characteristics of a target sound as best as possible, given some kind of similarity metric (or loss function)~\cite{horner1993machine,mitchell2007evolutionary}.

\textit{Iterative} sound-matching refers to works that mimic the back-and-forth process of listening to a synthesizer's output and updating the parameters until the termination conditions are met. The two necessary components are the \textit{loss function}, which determines how close two sounds are, and the \textit{parametric synthesizer} that outputs sounds based on input parameters. In the iterative workflow, a target sound is selected, the synthesizer outputs sounds based on its parameters, and the loss function determines the proximity of the output to the target. From there, the synthesizer parameters are updated, the output and target are compared, and so on.

Despite the early popularity of the iterative approach~\cite{justice1979analytic,horner1993machine,mitchell2007evolutionary}, recent works in sound-matching have favored inference models, deep neural-networks (\gls{DNN}), and large datasets. Rather than continually interacting with the environment, DNN-based models learn from a dataset of inputs (synthesizer parameters) and outputs (corresponding sound), and predict the synthesizer parameters for new sounds~\cite{yee2018automatic,han2023perceptual,han2024learning}.

We believe that iterative sound-matching is interesting for multiple reasons. First, it better resembles the human approach to sound-design, which requires continually interacting with the environment and making decisions according to a feedback signal~\cite{sutton2018reinforcement}. Second, the trial-and-error approach leaves a trail of parameter sets from the starting point to the end goal, which could be of use to sound designers. Third, relative to the DNN-based inference models, it provides a more tractable environment for exploring the complex issue of sound-similarity functions, and their interactions with different methods of synthesis.

With the goal of improving iterative sound-matching algorithms for sound designers, we identify a number of weaknesses in current literature (see Section~\ref{sec:lacking}). The experiments in this work pertain to two core issues, which we named \LossSelect~and \SynthSelect. The former refers to the search for an optimal loss function, lack of novel algorithms, and the---perhaps needless---focus on outperforming state of the art (\gls{SOTA}), while the latter refers to the lack of diversity in synthesis methods, and their interaction with the loss function.

@@

where $T$ represents time in seconds, and $n$ is the frequency parameter. If $n$ is 1, then $x(n)$ would be a waveform with a frequency of 1 hertz, meaning that the value of $x(n)$ oscillates to the same point once every second. Waveform generators are called oscillators for this reason. If $n$ is in the range of 30 to 20,000 (the upper limit is often lower for some), $x(n)$ could be played as an audible tone~\cite{smith2007mathematics,jilek2014reference}.

@@

Differential DSP (\gls{DDSP}) is a general term for applications that combine machine learning techniques such as gradient descent~\cite{goodfellow2016deep,boyd2004convex} with DSP~\cite{engel2020ddsp}. Implementing complex DSP functions in a differentiable manner can be challenging, and effective differentiable audio similarity measures require careful mathematical expression of the desired attributes of sound. These issues have likely contributed to the limited exploration of this domain. Perhaps to most relevant recent work is by Vahidi \textit{et al.}, featuring the comparison of two differentiable loss functions with a simple FM synthesizer~\cite{vahidi2023mesostructures}.

@@

In sound-matching, particularly in a Deep Learning (\gls{DL}) context~\cite{goodfellow2016deep}, the similarity function can also be called a \textit{loss} function, where the emphasis is on measurement and reduction of the distance between target and output. It is important to note that there is a close relationship between the loss function $L$ and the sound representation function $\phi$. $L$ is the result of a distance measure $d$ applied to the features extracted by $\phi$.

@@

\subsection{Automatic Sound Matching}
\label{sec:sound_matching_definition}
In past works the sound-matching problem has been tackled from a variety of perspectives as it involves a number of modular parts: the choice of synthesizer $g$, the target sounds of interest, the representation function $\phi$, and the heuristic for finding the optimal $\theta$. We discuss these components in the upcoming sections, and give a coherent typology of the sound-matching tasks in Section~\ref{sec:matching_types}.
